Building a continuous benchmarking ecosystem in bioinformatics

### 2024-09-23 manifesto-main.tex
funkyheatmaps

Benchmarks are often focused on finding the `top' performing method, or on generating ranked lists of methods. While this is valid in a few fields with clearly defined goals, such as image recognition \cite{Khan2018-ot} or protein structure prediction \cite{Jumper2021-oq}, bioinformatics tasks are typically evaluated by multiple metrics, some of which are not well understood \cite{Lutge2021-mt, Reinke2024-dw}. Additionally, ground truths are not always well-defined (e.g., derived with uncertainty from other experimental data), multiple tasks (e.g., represented by different reference datasets) may be evaluated for a single method and several ways to assess performance are possible. Even when accompanied by dataset- and/or metric-specific performance measures, scaling and aggregation of results make it difficult to unpack fine-grained nuances of performance from a final summary table reported in most benchmarks (e.g., \footnote{\url{https://www.nature.com/articles/s41592-021-01336-8/figures/11}}). Strobl \textit{et al.} push back against the `one method fits all datasets’ philosophy and advocate for highlighting dataset or parameter properties that are predictive of method performance \cite{Jelizarow2010-kr, Boulesteix2010-zj, Strobl2024-lp}. This aligns with what a data analyst wants: to identify the best performers in settings most similar to their problem at hand. This requires not only a thorough benchmark design, e.g., tracking of meaningful dataset metadata \cite{Strobl2024-lp}, but also can be supported by a benchmarking system that facilitates flexible navigation of results (e.g., via an interactive dashboard).

Besides dashboards for more nuanced result exploration \cite{MariniUnknown-aq}, multi-criteria decision analysis (MCDA) \cite{Taherdoost2023-wd} has been proposed to guide users through benchmark results and multidimensional scaling (MDS) has been used to differentiate the effect of single datasets or metrics and recognize patterns beyond aggregated performance \cite{Niessl2022-fk}. An extensive analysis and presentation of the results not only helps the users in the long term but also the whole scientific field to identify pain points, the complexity of possible scenarios and to isolate the different use-cases. The first advantage of such a strategy is to force the user to an attentive reading of the performance results so that the correct use-case and relevant methods can be identified. The second, more important, advantage to present benchmark results extensively is to identify the challenges in a field of analysis, i.e., the scenarios, types of datasets, or aspects of performance that are systematically associated with poor results for all current methods. Being public and open to contributions from specialists of different fields, a benchmarking system could gather more reference datasets and methods, and allow the current state of the field to be better grasped. %As a result, the performance results will tend to better fit to the current state of the field, as opposed to a static benchmark led by a small panel of experts that leads to a smaller, and sometimes (un)intentionally biased view of the field \cite{Boulesteix2010-zj, Weber2019-el}.

@@

A related challenge concerns which datasets and metrics to include in a benchmark. On the one hand, diverse sets of datasets and metrics are important to ensure generalizability and avoid overfitting. On the other hand, inclusion of low-quality or redundant datasets or metrics does not benefit the benchmark, and may provide misleading (aggregate) results. Possible mitigations may include sensitivity analyses, evaluating the impact of single datasets or metrics on the overall, aggregated performance of a method, or reducing redundancy by downweighting similar metrics to avoid a specific aspect dominating the evaluation. From a wider perspective, these questions are all related to the governance of a benchmark, a topic that has not received much attention. Ultimately benchmarks may want to, depending on their scale and complexity, implement different governance models. In one model, the benchmarkers have the ultimate authority to decide on what gets included in the benchmark, simplifying the decision making but creating a risk of gatekeeping. In another model, decisions are made collectively by a group of contributors, which may mitigate the risk of gatekeeping at least partially, but creates a more complex decision structure and introduces the need for a conflict resolution strategy. The governance model of a benchmark could also cover aspects related to computing infrastructure and storage, as well as guidelines to manage authorship of any resulting publications or resources. Having the governance model spelled out when launching a benchmark is likely to reduce friction at later stages of the project. In all cases, an advantage of doing things in public is that contributions are logged.

@@

Additionally, which level and scope of data validation needs to be decided. Validations include content integrity (e.g., md5sum), data syntax (e.g., JSON Schema), and data semantics (e.g., SHACL for linked data). Intermediate data and results should be made FAIR, leading to additional decisions, including the scope of data sharing (e.g., only the most important subset), the handling of the metadata (automatic versus manual annotation) and the provider (where and how to make available, e.g., GEO or Zenodo). Automatic generation of metadata might require predefined standards and an ontology to fully track how the data was generated \cite{LeboUnknown-om}.

>>>>

Benchmarks are often focused on finding the `top' performing method, or on generating ranked lists of methods. While this is valid in a few fields with clearly defined goals, such as image recognition \cite{Khan2018-ot} or protein structure prediction \cite{Jumper2021-oq}, bioinformatics tasks are typically evaluated by multiple metrics, some of which are not well understood \cite{Lutge2021-mt, Reinke2024-dw}. Additionally, ground truths are not always well-defined (e.g., derived with uncertainty from other experimental data), multiple tasks (e.g., represented by different reference datasets) may be evaluated for a single method and several ways to assess performance are possible. Even when accompanied by dataset- and/or metric-specific performance measures, scaling and aggregation of results make it difficult to unpack fine-grained nuances of performance from a final summary table reported as a FunkyHeatmap \cite{funkyheatmap} as in most benchmarks (e.g., \cite{Saelens2019-jy}). Strobl \textit{et al.} push back against the `one method fits all datasets’ philosophy and advocate for highlighting dataset or parameter properties that are predictive of method performance \cite{Jelizarow2010-kr, Boulesteix2010-zj, Strobl2024-lp}. This aligns with what a data analyst wants: to identify the best performers in settings most similar to their problem at hand. This requires not only a thorough benchmark design, e.g., tracking of meaningful dataset metadata \cite{Strobl2024-lp}, but also can be supported by a benchmarking system that facilitates flexible navigation of results (e.g., via an interactive dashboard).

Besides dashboards for more nuanced result exploration \cite{MariniUnknown-aq}, multi-criteria decision analysis (MCDA) \cite{Taherdoost2023-wd} has been proposed to guide users through benchmark results and multidimensional scaling (MDS) has been used to differentiate the effect of single datasets or metrics and recognize patterns beyond aggregated performance \cite{Niessl2022-fk}. An extensive analysis and presentation of the results not only helps the users in the long term but also the whole scientific field to identify pain points, the complexity of possible scenarios and to isolate the different use-cases. The first advantage of such a strategy is to force the user to an attentive reading of the performance results so that the correct use-case and relevant methods can be identified. The second, more important, advantage to present benchmark results extensively is to identify the challenges in a field of analysis, i.e., the scenarios, types of datasets, or aspects of performance that are systematically associated with poor results for all current methods. Being public and open to contributions from specialists of different fields, a benchmarking system could gather more reference datasets and methods, and allow the current state of the field to be better grasped.

%As a result, the performance results will tend to better fit to the current state of the field, as opposed to a static benchmark led by a small panel of experts that leads to a smaller, and sometimes (un)intentionally biased view of the field \cite{Boulesteix2010-zj, Weber2019-el}.

@@

A related challenge concerns which datasets and metrics to include in a benchmark. On the one hand, diverse sets of datasets and metrics are important to ensure generalizability and avoid overfitting. On the other hand, inclusion of low-quality or redundant datasets or metrics does not benefit the benchmark, and may provide misleading (aggregate) results. Possible mitigations may include sensitivity analyses, evaluating the impact of single datasets or metrics on the overall, aggregated performance of a method, or reducing redundancy by down-weighting similar metrics to avoid a specific aspect dominating the evaluation. From a wider perspective, these questions are all related to the governance of a benchmark, a topic that has not received much attention. Ultimately benchmarks may want to, depending on their scale and complexity, implement different governance models. In one model, the benchmarkers have the ultimate authority to decide on what gets included in the benchmark, simplifying the decision making but creating a risk of gatekeeping. In another model, decisions are made collectively by a group of contributors, which may mitigate the risk of gatekeeping at least partially, but creates a more complex decision structure and introduces the need for a conflict resolution strategy. The governance model of a benchmark could also cover aspects related to computing infrastructure and storage, as well as guidelines to manage authorship of any resulting publications or resources. Having the governance model spelled out when launching a benchmark is likely to reduce friction at later stages of the project. In all cases, an advantage of doing things in public is that contributions are logged.

@@

Additionally, which level and scope of data validation needs to be decided. Validations include content integrity (e.g., md5sum), adherence to a data specification (e.g., JSON Schema), and data semantics (e.g., SHACL for linked data). Intermediate data and results should be made FAIR, leading to additional decisions, including the scope of data sharing (e.g., only the most important subset), the handling of the metadata (automatic versus manual annotation) and the data storage provider and policies (where and how to make available, e.g., GEO or Zenodo). Automatic generation of metadata might require predefined standards and an ontology to fully track how the data was generated \cite{LeboUnknown-om}.
